Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-20 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: night revisions

\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[dvipsnames,table]{xcolor}
\usepackage{bm}
\usepackage{txfonts}
% \usepackage{newtxtext}
% \usepackage{newtxmath}
\usepackage{natbib}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{url}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{lineno}
\usepackage{mathrsfs}
\usepackage{makecell}
\usepackage{microtype}

@@

% Comment commands
\usepackage{xcolor}
\newcommand{\agcomm}[1]{{\textcolor{red}{{[AG: #1]}} }}
\newcommand{\ag}[1]{{\textcolor{Maroon}{{#1}} }}
\newcommand{\hs}[1]{{\textcolor{blue}{{[HS: #1]}} }}
\newcommand{\ab}[1]{{\textcolor{cyan}{{#1}} }}
\newcommand{\abcomm}[1]{{\textcolor{red}{{[AB: #1]}} }}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
\ab{Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is}
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we \sout{take a first step to} address this question by using an
inspiral-merger-ringdown waveform model in the effective-one-body formalism,
augmented with a parametrization of the ringdown based on an expansion \ab{in} the
final black hole's spin.
%
We give a prescription on how to include \ab{in} this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. \ab{In particular, we focus on gravity} theories that deviate
from general relativity through higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against \ab{the two loudest gravitational-wave ringdown signals, so far,
  notably, GW150914 and GW200129. We find} that while the Einstein-dilaton-Gauss-Bonnet theory cannot be constrained with
these events, we can place upper bounds on the coupling constants of cubic-
($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order ($\ell_{\rm qEFT}
\leqslant 51.3$~km) effective-field-theories of general relativity, and
of dynamical Chern-Simons gravity ($\ell_{\rm CS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently
unconstrained by inspiral-only analyses which, however, can be constrained by
merger-ringdown studies \ab{with current gravitational-wave data}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where \ab{$g \equiv \textrm{det}(g_{\mu \nu})$} is the metric determinant, $R$ is the Ricci
scalar, $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
%
and $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and can be divided into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) \equiv \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solutions as GR and those of scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.
BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $r^{-1}$ fall-off of the scalar field, \ab{where $r$ is the distance from the BH.}
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is sourced by a curvature scalar, $Q$ is larger (smaller),
\ab{the} smaller (larger) the BH mass is.

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant \ab{quadrupolar} GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least \ab{if} one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron-star--black-hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of binary BHs (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}~\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
Secondly, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational \ab{perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.}
%

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ by itself is a boundary
term~\cite{Jackiw:2003pm} that does not contribute to the field equations in
four-dimensions.
%
However, a coupling with $\vartheta$ can modify the field equations.
%
The coupling strength is set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits as a solution, the garden-variety Schwarzschild BH of GR. This
is not the case when rotation is included and the Kerr metric is not a solution of
the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg}.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order~\cite{Yagi:2011xp,Alexander:2017jmt}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
\ab{only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.XXX of
Ref.~\cite{LIGOScientific:2021sio}). Thus, dCS gravity is poorly constrained by
analyses that rely on the inspiral portion of GWs when using current BBH observations are used~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated neutron star
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}.

The QNMs of the Schwarzschild BH in dCS gravity were studied in
Refs.~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb}, which found that scalar
perturbations couple to gravitational perturbations of axial parity, \ab{in contrast
with EdGB gravity.}
%
This leads to a breakdown of isospectrality. The QNM spectra of slowly-rotating BHs in dCS gravity was
studied in Ref.~\cite{Wagle:2021tam}.
%
They were also extracted from numerical-relativity simulations of head-on BH collisions
in Ref.~\cite{Okounkova:2019dfo}.

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell_{\rm EFT}^{2n - 2}\, L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell_{\rm EFT}$ is a length-scale assumed to be small compared to the \ab{length-scale $M$
associated with a BH (i.e., $\ell_{\rm EFT} / M \ll 1$),} and $L^{(2n)}$ are corrections
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and \sout{work up} \ab{consider up} to dimension-eight operators
($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and $(\lambda_{\rm o, e},\varepsilon_{i})$,  $i=1,2,3$, are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we focus on a subset of the parameter space.
%
In particular, we consider dimension-six and dimension-eight operators separately.
%
\ab{For simplicity,} in the dimension-six case we further assume that $\lame = \lamo = 1$, leaving us with $\ell = \ell_{\rm cEFT}$
as our single free parameter.
%
\ab{Similarly,} in the dimension-eight case, we set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell = \ell_{\rm qEFT}$ as our single free parameter.

In general \dots
%
\hs{Alessandra, can you include a short description of~\cite{Sennett:2019bpc} for the quartic EFT?}

@@

\ab{Having reviewed the modified gravity theories that we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the parametrized ringdown spin-expansion coefficients (ParSpec) framework in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral(--plunge)-merger-ringdown (IMR) waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv}, and explain how we extend
it to include the ParSpec.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
free coefficients in the ParSpec framework.}
%
Ultimately, this provides us with an IMR waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

\ab{A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of the GR BH is to write,}
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
\ab{where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.}
%
This type of parametrization was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

The current LIGO-Virgo-KAGRA tests of BH ringdown (see Sec.VII.A
of Ref.~\cite{LIGOScientific:2020tif} or Sec.VIII.A
of Ref.~\cite{LIGOScientific:2021sio}) take a restrictive \ab{theory-independent}
approach towards the inference of $\delta\omega_{\ell m n}$ and
$\delta\tau_{\ell m n}$. These deviations are either assumed to occur
identically across all observed sources or belong to a generic
underlying Gaussian population. A more realistic scenario, however, is
that these parameters depend on the source BH's mass and spin.
%
Ideally, one would like to explicitly reinstate this dependence, by
%
(i) introducing deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) making it simpler to combine constraints coming from multiple (independently observed) sources.

>>>>

\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[dvipsnames,table]{xcolor}
\usepackage{bm}
\usepackage{txfonts}
% \usepackage{newtxtext}
% \usepackage{newtxmath}
\usepackage{natbib}
\usepackage{amssymb}
\usepackage{xspace}

\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\newcommand\redsout{\bgroup\markoverwith{\textcolor{red}{\rule[0.5ex]{8pt}{1pt}}}\ULon} % To get a red strikethrough
% (which is easier to spot)

\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{url}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{lineno}
\usepackage{mathrsfs}
\usepackage{makecell}
\usepackage{microtype}

@@

% Comment commands
\usepackage{xcolor}
\newcommand{\agcomm}[1]{{\textcolor{red}{{[AG: #1]}}}}
\newcommand{\ag}[1]{{\textcolor{Maroon}{{#1}}}}
\newcommand{\hs}[1]{{\textcolor{TealBlue}{{#1}}}}
\newcommand{\hscomm}[1]{{\textcolor{TealBlue}{{[HS: #1]}}}}
\newcommand{\ab}[1]{{\textcolor{cyan}{{#1}}}}
\newcommand{\abcomm}[1]{{\textcolor{red}{{[AB: #1]}}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% General:
The detection of gravitational waves produced by coalescing compact binaries
allows us to explore the dynamical, nonlinear regime of general relativity
and to constrain modifications to it.
%
Some of these gravitational-wave events have sufficiently high
signal-to-noise ratio in the merger, allowing us to probe the relaxation of the remnant black
hole to its final, stationary state --- the so-called black-hole ringdown, which is
characterized by a set of quasinormal modes.
% Main question:
Can we use the ringdown to constrain deviations to general relativity,
as predicted by several of its contenders?
% What we do:
Here, we  address this question by using an inspiral-merger-ringdown waveform
model in the effective-one-body formalism, augmented with a parametrization of
the ringdown based on an expansion in the final black hole's spin.
%
We give a prescription on how to include in this waveform model the quasinormal mode frequencies
calculated on a theory-by-theory basis. In particular, we focus on gravity theories that deviate
from general relativity through higher-order curvature corrections, namely,
Einstein-dilaton-Gauss-Bonnet, dynamical Chern-Simons, cubic- and quartic-order
effective-field-theories of general relativity.
%
% Main conclusion:
We confront the resultant waveform models against the two loudest
gravitational-wave ringdown signals, so far, notably, GW150914 and GW200129.
%
We find that while the Einstein-dilaton-Gauss-Bonnet theory cannot be
constrained with these events, we can place upper bounds on the coupling
constants of cubic- ($\ell_{\rm cEFT} \leqslant 38.2$~km) and quartic-order
($\ell_{\rm qEFT} \leqslant 51.3$~km) effective-field-theories of general
relativity, and of dynamical Chern-Simons gravity ($\ell_{\rm dCS} \leqslant 38.7$~km).
%
The latter result is a concrete example of a theory presently unconstrained by
inspiral-only analyses which, however, can be constrained by merger-ringdown
studies with current gravitational-wave data.
%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%

@@

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm EdGB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm EdGB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g \equiv \textrm{det}(g_{\mu \nu})$ is the metric determinant, $R$ is the Ricci
scalar, $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
%
and $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and can be divided into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi)
\equiv \dd f  / \dd \varphi$ is always nonzero and BHs are known to always
support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which
$f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solutions as GR and those of scalarized
BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.
BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $r^{-1}$ fall-off of the scalar field, where $r$ is the distance from the BH.
%
This charge is not an independent parameter, and depends on the BH's mass and spin,
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is sourced by a curvature scalar, $Q$ is larger (smaller),
the smaller (larger) the BH mass is.

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least if one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 1.33$~km
%
using the neutron-star--black-hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of binary BHs (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}~\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
Secondly, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.
%

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{align} \label{eq:action_dcs}
    S_{\rm dCS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm dCS} \, \vartheta \, {}^{*}RR
    \right],
    \nonumber \\
\end{align}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ by itself is a boundary
term~\cite{Jackiw:2003pm} that does not contribute to the field equations in
four-dimensions.
%
However, a coupling with $\vartheta$ can modify the field equations.
%
The coupling strength is set by the coupling constant $\ell_{\rm dCS}$, with
dimensions of length.

The theory admits as a solution, the garden-variety Schwarzschild BH of GR. This
is not the case when rotation is included and the Kerr metric is not a solution of
the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg}.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order~\cite{Yagi:2011xp,Alexander:2017jmt}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
\ab{only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.XXX of
Ref.~\cite{LIGOScientific:2021sio}). Thus, dCS gravity is poorly constrained by
analyses that rely on the inspiral portion of GWs when using current BBH observations are used~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated neutron star
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}.

The QNMs of the Schwarzschild BH in dCS gravity were studied in
Refs.~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb}, which found that scalar
perturbations couple to gravitational perturbations of axial parity, in contrast
with EdGB gravity.
%
This leads to a breakdown of isospectrality. The QNM spectra of slowly-rotating BHs in dCS gravity was
studied in Ref.~\cite{Srivastava:2021imr,Wagle:2021tam}.
%
They were also extracted from numerical-relativity simulations of head-on BH collisions
in Ref.~\cite{Okounkova:2019dfo}.

@@

Our last example of modified gravity theories are the so-called effective field
theories (EFT) of GR~\cite{Endlich:2017tqa,Sennett:2019bpc,deRham:2020ejn,Cano:2020cao,Cano:2021myl}.
%
They are described by the action
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \sum_{n \geqslant 2} \ell_{\rm EFT}^{2n - 2}\, L^{(2n)}
    \right]\,,
\end{equation}
%
where $\ell_{\rm EFT}$ is a length-scale assumed to be small compared to the length-scale $M$
associated with a BH (i.e., $\ell_{\rm EFT} / M \ll 1$), and $L^{(2n)}$ are corrections
that introduce higher-order curvature tensors (with $2n$ metric derivatives).

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and consider up to dimension-eight
operators ($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and $(\lambda_{\rm o, e},\varepsilon_{i})$, $i=$~1, 2, 3, are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we focus on a subset of the parameter space.
%
In particular, we consider dimension-six and dimension-eight operators separately.
%
In addition, in the dimension-six case we further assume that $\lame = \lamo = 1$,
leaving us with \redsout{$\ell =$} $\ell_{\rm cEFT}$ as our single free parameter.
%
Similarly, in the dimension-eight case, we set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with \redsout{$\ell =$} $\ell_{\rm qEFT}$ as our single free parameter.

In general \dots
%
\hscomm{Alessandra, can you include a short description of~\cite{Sennett:2019bpc} for the quartic EFT?}

@@

Having reviewed the modified gravity theories that we will consider, we now present
the sequential building blocks for the waveform model we use to test these theories
against GW observations.
%
We start by reviewing the parametrized ringdown spin-expansion coefficients
(ParSpec) framework~\cite{Maselli:2019mjd} in Sec.~\ref{sec:review_parspec}.
%
Next, in Sec.~\ref{sec:review_pSEOB}, we review our baseline
parametrized inspiral(--plunge)-merger-ringdown (IMR) waveform
model~\cite{Brito:2018rfr,Ghosh:2021mrv}, and explain how we extend
it to include the ParSpec.
%
Finally, in Sec.~\ref{sec:theory_specific_qnm}, we show how we can map
theory-specific QNM calculations in modified gravity theories onto the
free coefficients in the ParSpec framework.
%
Ultimately, this provides us with an IMR waveform model, with the ringdown
portion of the model informed by QNM calculations in specific beyond-GR
theories.

@@

A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of BHs of GR is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization\footnote{\hs{See also
Refs.~\cite{Glampedakis:2017dvb,Glampedakis:2019dqh,Silva:2019scu} and
Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj,Kimura:2020mrh} for alternative parametrizations.}}
was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

The current LIGO-Virgo-KAGRA tests of BH ringdown (see Sec.VII.A
of Ref.~\cite{LIGOScientific:2020tif} or Sec.VIII.A
of Ref.~\cite{LIGOScientific:2021sio}) take a restrictive theory-independent
approach towards the inference of $\delta\omega_{\ell m n}$ and
$\delta\tau_{\ell m n}$. These deviations are either assumed to occur
identically across all observed sources or belong to a generic
underlying Gaussian population. A more realistic scenario, however, is
that these parameters depend on the source BH's mass and spin.
%
Ideally, one would like to explicitly reinstate this dependence, by
%
(i) introducing deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) making it simpler to combine constraints coming from multiple (independently observed) sources.
